Classless Java: Tuning Java Interfaces

### 2016-03-09 OOPSLA16/sections/Introduction.tex

Object oriented languages always strive to offer great code reuse,
they aim to couple flexibility and rigour, expressive power and modular reasoning.
Two main idea have emerged to this end: prototype based  and class based language.
In prototype based language an objects inherits from other objects,
and thus objects own both behaviour and state (and objects is all you have).
In class based languages an object is instance of a specific class, and classes inherits from other classes.
Here objects own the state, while classes contains behaviour and the structure of the state.

We present here a third alternative: the concept of
 \textbf{Interface based object oriented languages}, where 
  objects implements interfaces, and they own implementation for the behaviour,
that is structurally defined in their interface.

@@

  The literature of traits provide good examples on how easy and modular can be combining
  multiple sources of behaviour, and the literature over multiple inheritance show us how
  hard it is to modularly combine multiple sources of \textbf{and} state.
  However, our object are the only responsible to define 
  the ultimate behaviour or a method, and if such method is just a getter/setter.

We could define our approach as a novel language, with new syntax and semantic, but this would have a
steep learning curve; so for the sake of providing a more accessible explanation, we will encode our ideas
over Java, relying on the new features of Java 8; namely
interface \emph{static methods}, and \emph{default methods}, allowing interfaces to have
method implementations. While the motivation for default methods was
to allow interfaces to be extended over time while preserving backwards
compatibility (\emph{interface evolution}),
 default methods can also emulate \emph{traits}~\cite{scharli03traits}.
\footnote{The
original notion of traits by Scharli et al. prescribes, among other
things, that: 1) a trait provides a set of methods that implement
behavior; and 2) a trait does not specify any state variables, so the
methods provided by traits do not access state variables
directly. Java 8 interfaces follow similar principles too. Indeed, a
detailed description of how to emulate trait-oriented programming in
Java 8 can be found in the work by Bono et al.~\cite{bono14}. The Java 8
team designing default methods, was also fully aware of that secondary
use of interfaces, but it was not their objective to model traits:
``The key goal of adding default methods to Java was "interface
evolution", not "poor man's traits"''~\cite{goetz13default}. As a result, 
they were happy to support the secondary use of interfaces with
default methods as long as it did not make the implementation and
language more complex.}

@@

We propose a programming style, where we never use classes\footnote{More precisely, we never use the \Q@class@ keyword.}.
We call this restricted version of Java \emph{class-less java}.
Our proposed style enables powerful object-oriented idioms, similar to 
multiple inheritance.

Since Java was not designed to be used in this way, our style can get pretty verbose, especially about
object instantiation.
We will annotate with \Q|@Obj| certain interfaces and rely on
the Annotation processing to generate boring and repetitive code
allowing interface instantiation.
We call such annotated interface \emph{object interfaces};
they support \emph{abstract state operations}, providing a way
to use multiple inheritance with state in Java. The abstract state
operations include various common utility methods (such as getters and
setters, or clone-like methods). In the presence of subtyping, such
operations often require special care, as their types need to be
refined. Object interfaces provide support for type-refinement and can
automatically produce code that deals with type-refinement
adequately. %With object interfaces, many Java programs can be built
%without using a single class!

>>>>

Object oriented languages always strive to offer great code reuse,
they aim to couple flexibility and rigour, expressive power and modular reasoning.
Two main idea have emerged to this end: prototype based (PB) and class based (CB) languages.
In prototype based languages an objects inherits from other objects,
and thus objects own both behaviour and state (and objects is all you have).
In class based languages an object is instance of a specific class, and classes inherits from other classes.
Here objects own the state, while classes contains behaviour and the structure of the state.

We present here a third alternative: the concept of
 \textbf{Interface based object oriented languages} (IB), where 
  objects implements interfaces, and they own implementation for the behaviour,
that is structurally defined in their interface.

@@

  The literature provide good examples on how easy and modular can be combining
  multiple sources of pure behaviour (traits[]), and how
  hard it is to modularly combine multiple sources of behaviour \textbf{and} state (multiple inheritance[]).
  However, our object are the only responsible to define 
  the ultimate behaviour or a method, and if such method is just a getter/setter.

Our idea could be explained by defining a novel language, with new syntax and semantic.
However, this would have a
steep learning curve; for the sake of providing a more accessible explanation, we will encode
\footnote{
In the same sense, you can encode prototype based programming over class based with
the strategy pattern, and class based programming with prototypes by creating ``class'' objects.
} our ideas
over Java, relying on the new features of Java 8; namely
interface \emph{static methods}, and \emph{default methods}, allowing interfaces to have
method implementations. While the motivation for default methods was
to allow interfaces to be extended over time while preserving backwards
compatibility (\emph{interface evolution}),
 default methods can also emulate \emph{traits}~\cite{scharli03traits}.
\footnote{The
original notion of traits by Scharli et al. prescribes, among other
things, that: 1) a trait provides a set of methods that implement
behavior; and 2) a trait does not specify any state variables, so the
methods provided by traits do not access state variables
directly. Java 8 interfaces follow similar principles too. Indeed, a
detailed description of how to emulate trait-oriented programming in
Java 8 can be found in the work by Bono et al.~\cite{bono14}. The Java 8
team designing default methods, was also fully aware of that secondary
use of interfaces, but it was not their objective to model traits:
``The key goal of adding default methods to Java was "interface
evolution", not "poor man's traits"''~\cite{goetz13default}. As a result, 
they were happy to support the secondary use of interfaces with
default methods as long as it did not make the implementation and
language more complex.}

@@

We propose a programming style, where we never use classes\footnote{More precisely, we never use the \Q@class@ keyword.}.
We call this restricted version of Java \emph{Class-less Java}.
Our proposed style enables powerful object-oriented idioms, similar to 
multiple inheritance.

Since Java was not designed to be used in this way, our style can be verbose, especially about
object instantiation.
We will annotate with \Q|@Obj| certain interfaces and rely on
the Annotation processing to generate boring and repetitive code
allowing interface instantiation.
We call such annotated interface \emph{object interfaces};
they support \emph{abstract state operations}, providing a way
to use multiple inheritance with state in Java. The abstract state
operations include various common utility methods (such as getters and
setters, or clone-like methods). In the presence of subtyping, such
operations often require special care, as their types need to be
refined. Object interfaces provide support for type-refinement and can
automatically produce code that deals with type-refinement
adequately. %With object interfaces, many Java programs can be built
%without using a single class!

### 2016-03-09 OOPSLA16/sections/Overview.tex

The animal system modeled so far is a simplified version of the
system presented in the left-side of Figure~\ref{fig:pegasus}, which still does not appear to
justify the \mixinAnn{} annotation.  Now moving on to modeling
the complete animal system where \Q@Animal@s will include a \Q@location@
representing their positions in space, we shall
see that modeling stateful components using plain Java 8 quickly becomes
cumbersome. The \mixinAnn{} annotation comes to rescue here and
 avoids significant amounts of boilerplate code.

\paragraph{\Q@Point2D@: simple immutable data with fields}
2D points should keep track of their coordinates.  The usual
approach to model points in Java would be to use a class with fields for the
coordinates. However here we will illustrate how points are modeled with
interfaces. Since Java disallows fields inside interfaces, state is modeled
using abstract (getter) methods:

@@

\noindent Unfortunately, creating a new point object is cumbersome, even
with anonymous classes:

@@

\noindent However this cumbersome syntax is not required for every
object allocation. As programmers do, for ease or reuse, the boring
repetitive code can be encapsulated in a method. A generalization of the
\texttt{of} static factory method is appropriate in this case:
\begin{lstlisting}
interface Point2D { int x(); int y();
    static Point2D of(int x, int y) { return new Point2D() {
        public int x(){return x;}
        public int y(){return y;}};}
}
\end{lstlisting}

@@

\noindent Only code for methods needing implementation is generated. Thus,
programmers can easily customize the behavior for their special needs.
%\end{comment}

@@

\caption{Abstract state operations, for a field \texttt{x}, allowed by the \mixin
  annotation. }

@@

\subsection{Object Interfaces and Subtyping}
\Q@Bird@s are \Q@Animal@s, but while \Q@Animal@s only need 2D
locations, \Q@Bird@s need 3D locations. Therefore when the \texttt{Bird}
interface extends the \Q@Animal@ interface, the notion of points needs to
be refined. Such kind of refinement usually poses a challenge
in typical class-based approaches. Fortunately, with object interfaces,
we are able to provide a simple and effective solution.

@@

\noindent From the type perspective, the key is the covariant method
overriding of \texttt{location()}. However, from the semantic
perspective the key is the implementation for the setter with the old
signature (\Q@location(Point2D)@). The key to the setter
implementation is a new type of \Q@with@ method, called a property updater.

@@

>>>>

The animal system modeled so far is a simplified version of the
system presented in the left-side of Figure~\ref{fig:pegasus}.
Here you still can not appreciate the advantage of IB.
Lets model now the complete animal system where \Q@Animal@s will include a \Q@location@
representing their positions in space. We will use 2D points to keep track of locations using coordinates.

\paragraph{\Q@Point2D@: simple immutable data with fields}
Here we will illustrate how points are modelled with
interfaces. In IB we do not talk about state directly; state is accessed
using abstract (getter) methods. Note how in   The usual
approach to model points in Java would be to use a class with fields for the
coordinates.

@@

\noindent Our encoding over Java is now inconvenient: creating a new point object is cumbersome, even
with anonymous classes:

@@

\noindent However this cumbersome syntax is not required for every
object allocation. As programmers do, for ease or reuse, the boring
repetitive code can be encapsulated in a method. A generalization of the
\texttt{of} static factory method is appropriate in this case:
\begin{lstlisting}
interface Point2D { int x(); int y();
  static Point2D of(int x, int y) {
    return new Point2D() {
      public int x(){return x;}
      public int y(){return y;}
    };  }  }
\end{lstlisting}

@@

\noindent Only code for methods needing implementation is generated. Thus,
programmers can easily customize the behaviour for their special needs.
Also, since \mixin interfaces offer the \Q@of@ factory method, only interfaces where all the abstract methods
can be synthesized can be \mixin. A non \mixin interface is like an abstract class in Java.
%\end{comment}

@@

\caption{Abstract state operations, for a field \texttt{x}, allowed by the \mixin
  annotation. \marco{I disagree with this table, for example we miss the with} }

@@

\subsection{Object Interfaces and Subtyping}
\Q@Bird@s are \Q@Animal@s, but while \Q@Animal@s only need 2D
locations, \Q@Bird@s need 3D locations. Therefore when the \texttt{Bird}
interface extends the \Q@Animal@ interface, the notion of points needs to
be refined. Such kind of refinement is challenging
in typical class-based approaches. Fortunately, with object interfaces,
we are able to provide a simple and effective solution.

@@

\noindent From the type perspective, the key is the covariant method
overriding of \texttt{location()}. However, from the semantics
perspective the key is the implementation for the setter with the old
signature (\Q@location(Point2D)@). The key to the setter
implementation is a new type of \Q@with@ method, called a property updater.

@@

\subsection{Abstract state operations}
Defining an convenient list of abstract state operation is not an easy task.
Of course we need getters.
Currently we are offering both setters and Fluent setters. May be fluent setters are enough, and we can drop the
concept of setters?
We offer with withers and the with.
What can be a nice syntax?
\marco{my coments from now on}
A language natively supporting Object interfaces may offer the following syntax:
\begin{lstlisting}
obj interface Foo{
  int bar;//just getter
  var int bar;//also setter(fluent)
  with int bar;//also wither
  with var int bar;//getter setter wither
  with/var int bar=2;//default value, not asked in of
  invariant{/*stuff that should always hold*/}
  updater Bar, Beer//multi setter
  wither Foo, Beer//multi with
  factory of(...){call generate of}// generated of is now private
}
\end{lstlisting}
